---
title: "Maroubra Bay Analysis"
output:
  html_document:
    df_print: paged
Course: BEES2041 Data Analysis for Life and Earth Sciences
Edition: Prime
Date: Dec 25, 2021
---
# Objectives

Maroubra Bay is a popular beach situated in Sydney, NSW, Australia. Over the years, the BEES2041 course cohort has collected multiple species composition data of the intertidal community, recorded in Maroubra_rock_pools.csv. The 2021 cohort also collected gastropods abundance data based on the height of the sample zones, in MaroubraZones.csv.

This project aims to analyse the aforementioned datasets to answer the following questions:

  1. Does mean abundance of gastropods vary across zones and species at Maroubra bay?
  2. How does gastropods species composition distribute in rockpool vs emergent habitats, and among sampling years?

The following section and the remainder of this notebook details the code from analysis. For detailed information on data collection, background introduction of the project, as well as further interpretation of the results, see PDF in the repository.

# Analysis
## Part A. Data Import
```{r}
library(tidyverse)
library(vegan)
```

```{r}
# Gastropods abundance with zone (defined by height on shore) data from Spring 2021, sampled from Maroubra Bay, NSW
Gastropods <- read.csv(file = "MaroubraZones.csv")

# Previous year (cohort 2016-2020, inclusive) data on multiple species composition at Maroubra Bay, NSW
RockPools <- read.csv(file = "Maroubra_rock_pools.csv", header = TRUE)
```

## Part B. Approaching the Questions
### 1. Does mean abundance of gastropods vary across zones and species at Maroubra bay?
```{r}
Gastropods$Zone <- factor(Gastropods$Zone, levels = c("Mid", "High")) ## stops ordering in alphabet. Instead, in our designated order

# Vis Gastropods dataset
ggplot(Gastropods, aes(Species, Abundance, fill=Zone)) + geom_boxplot() + theme_bw()
```
#### i. Bar plot: Abundance per zone
```{r}
Gastropods %>%
  group_by(Species, Zone) %>%
    summarise(mean.A = mean(Abundance),
    sd.A = sd(Abundance),
    SE.A = sd(Abundance)/sqrt(n())) %>%

# Bar plot
  ggplot(aes(Zone, mean.A, fill=Species)) +
    geom_bar(stat = "identity", position = position_dodge()) +

    # [optional] geom_text(aes(label = paste(round(mean.A,2))), position = position_dodge(0.9), vjust =-0.6, hjust=-0.6, size = 3.5) + ## labels mean data on top of bar

    geom_errorbar(aes(ymin = mean.A-SE.A, ymax = mean.A+SE.A), width = 0.3, position = position_dodge(0.9)) +
    ylab("Mean count per quadrat ± SE") +
    theme_classic() -> Fig_1

ggsave("bar_plot_maroubra_gastropods.png", width = 6, height = 4)

Fig_1
```
#### ii. 2-ANOVA: Abundance among zones and gastropod species
```{r}
# 2-ANOVA
Gastropods.ANOVA <- aov(Abundance ~ Zone * Species, data = Gastropods)
# Residual diagnostics
hist(Gastropods.ANOVA$residuals)
plot(Gastropods.ANOVA, which=c(1,2,3))
```

From residual diagnostics we see that HOV is not established. Log and Sqrt transforms were deployed yet did not address the HOV issue.
From the practical:

> You have heterogeneous variances and are violating one of the assumptions of ANOVA. This often happens with data that samples patchily distributed organisms.
  You cannot solve the problem but can consider its impact on the ANOVA; It may lead to an *artificially inflated F ratio*. Therefore proceed to run the ANOVA using the abundance data and **only reject null hypotheses if P < 0.01**. This is one way to address the possibility that you’d falsely reject your null hypothesis just because of heterogeneous variances.

We proceed the ANOVA using the original abundance data with an adjusted $\alpha$=0.01 as heterogenous variance was detected.

```{r}
summary(Gastropods.ANOVA)
```

Significant interaction between factors Zone:Species detected:
```{r}
# [optional] interaction.plot(Gastropods$Zone, Gastropods$Species, Gastropods$Abundance)

# Better plot:
Gastropods$Zone <- factor(Gastropods$Zone, levels = c("Mid", "High")) # Mid -> High, consistent with the bar plot

Gastropods %>%
  group_by(Species, Zone) %>%
    summarise(mean.A = mean(Abundance)) %>%

  ggplot(aes(Zone, mean.A, color=Species)) + geom_point() +
    geom_line(aes(y = mean.A, group = Species), size=1) +
    ylab("Mean abundance (count per quadrat)") +
    theme_bw() +
    theme(aspect.ratio = 1/2) -> Fig_2

ggsave("interaction_plot_gastropods.png", width = 10, height = 5)

Fig_2
```

Tukey's post-hoc:
```{r}
TukeyHSD(Gastropods.ANOVA)
```

### 2. (a) How does gastropods species composition distribute in rockpool vs emergent habitats, and (b) among sampling years?

First, inspect the dataset:
```{r}
view(RockPools)
```

Aim: Visualise species composition (reflected by individual species abundance) across habitats and sampling years.

#### i. MDS by (1) habitat (2) sampling year
```{r}
# MDS
RockPools_vars <- select(RockPools, -Habitat, - Year, -Replicate) ## Removing non-abundance data to form the key-variable matrix
RockPools_vars_std <- decostand(RockPools_vars, method= "max") ## Standardisation*
RP.sqrt <- sqrt(RockPools_vars_std + 1) ## Count data --> SQRT transform
```

*Algae abundance recorded in %cover, while gastropods and others recorded in counts. method="max" standardised each variable using respective maximum.

```{r}
# Creating the MDS profile
RockPoolsSqrt.S <- metaMDS(RP.sqrt, distance = "bray", autotransform = FALSE, trace=FALSE)

## This generates a similarity matrix with measure = Bray-Curtis similarity transform. Basis = standardised multivariable mtx we created previously.
## SQRT reduces MDS stress.
## Bray-Curtis similarity coefficient: Recommended for analyses that contrast the species composition of community data sets.


# Preparing the data frame for the MDS plot
RPsqrt_xy <- data.frame(RockPoolsSqrt.S$points) ## Sending the coordinates in MDS profile to new df

RPsqrt_xy$Habitat <- RockPools$Habitat ## Habitat profile "fill-in"
RPsqrt_xy$Year <- RockPools$Year ## Year profile "fill-in"
```

Now, we can generate the respective MDS plots using ggplot2:

```{r}
# MDS: (1) By habitat
Fig_3a <- ggplot(RPsqrt_xy, aes(MDS1, MDS2, color = Habitat)) + geom_point() +
  theme(axis.title = element_blank(),
        panel.background = element_rect(fill = "grey22", colour = "black", size = 1),
        panel.grid.major = element_line(color = "grey30", size=0.1),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 11), legend.title = element_text(size = 14),
        legend.key = element_rect(fill = NA)) +
  guides(color = guide_legend(override.aes = list(size = 4.5)))

ggsave("MDS_habitat.png", width = 10, height = 7)

Fig_3a

# MDS: (2) By sampling years
Fig_3b <- ggplot(RPsqrt_xy, aes(MDS1, MDS2)) + geom_point(aes(color = Year)) +
  theme(legend.key = element_rect(fill = NA),
        legend.text = element_text(size = 11),legend.title = element_text(size = 14),
        axis.title = element_blank(),
        panel.background = element_rect(fill = "grey97", colour = "black", size = 1),
        panel.grid.major = element_line(color = "grey86", size=0.1),
        panel.grid.minor = element_blank(),
        aspect.ratio = 7 / 9.9) +
  guides(color = guide_legend(override.aes = list(size = 3)))

ggsave("MDS_sampling_year.png", width = 10, height = 7)

Fig_3b

RockPoolsSqrt.S$stress ## Checking MDS stress for interpretation
```
#### ii. PERMANOVA by (1) habitat (2) sampling year
```{r}
# PERMANOVA: (1) By habitat
RPHabitat.adonis <- adonis(RockPools_vars_std ~ RockPools$Habitat, permutations = 999, method = "bray")
adonis(RockPools_vars_std ~ RockPools$Habitat, permutations = 999, method = "bray")
densityplot(permustats(RPHabitat.adonis))


# PERMANOVA: (2) By sampling years
RPYear.adonis <- adonis(RockPools_vars_std ~ RockPools$Year, permutations = 999, method = "bray")
adonis(RockPools_vars_std ~ RockPools$Year, permutations = 999, method = "bray")
densityplot(permustats(RPYear.adonis))
```
#### iii. Dispersion Analysis
```{r}
# 1. Defining group distance based on Bray-Curtis
# Recall RockPools --variable selection-> RockPools_vars --standardised via method="max"-> RockPools_vars_std --sqrt transform-> RP.sqrt
# i.e. RP.sqrt contains multiple abundance data that were standardised and transformed by sqrt(x+1)

vegdist(RP.sqrt, method="bray") -> RP.Bray  ## Maps aforementioned data using Bray-Curtis
```
```{r}
# 2. [OPTIONAL] Model using group medians with meta-analyses

# 2-1. Dispersion by habitat

# 2-1-1. Calculate multivariate dispersions
RPH.Bray <- betadisper(RP.Bray, RockPools$Habitat)
RPH.Bray

# 2-1-2. Perform test
anova(RPH.Bray)

# 2-1-3. Permutation test for F
permutest(RPH.Bray, pairwise = TRUE, permutations = 99)

# 2-1-4. Tukey's Honest Significant Differences
TukeyHSD(RPH.Bray)

# 2-1-5. Plot the groups and distances to centroids on the first two PCoA axes
plot(RPH.Bray)

# 2-1-6. [OPTIONAL] Draw a box plot of the distances to centroid for each group
# boxplot(RPH.Bray)

# 2-1-7. [OPTIONAL STATS] Group centroids/medians
# scores(RPH.Bray, 1:4, display = "centroids")

# 2-1-8. [OPTIONAL STATS] Eigenvalues from the underlying principal coordinates analysis
# eigenvals(RPH.Bray)
```
```{r}
# 2-2. Dispersion by sampling year

# 2-2-1. Calculate multivariate dispersions
RPY.Bray <- betadisper(RP.Bray, RockPools$Year)
RPY.Bray

# 2-2-2. Perform test
anova(RPY.Bray)

# 2-2-3. Permutation test for F
permutest(RPY.Bray, pairwise = TRUE, permutations = 99)

# 2-2-4. Tukey's Honest Significant Differences
TukeyHSD(RPY.Bray)

# 2-2-5. Plot the groups and distances to centroids on the first two PCoA axes
plot(RPY.Bray)
```

```{r, warning = False}
# 3. Group median with bias correction
# 3-1. By habitat
RPH.corr <- betadisper(RP.Bray, RockPools$Habitat, type = "median", bias.adjust=TRUE)

# 3-1-1. Coordinate mapping
RPH.corr.crs <- data.frame(RPH.corr$vectors)

RPH.corr.crs$Habitat <- RockPools$Habitat  ## Habitat column "fill-in"

# 3-1-2. ggplot: PCA with Habitat coloured
ggplot(RPH.corr.crs, aes(x=PCoA1, y=PCoA2, color=Habitat)) + geom_point() +
  theme(axis.title = element_blank(),
        panel.background = element_rect(fill = "grey22", colour = "black", size = 1),
        panel.grid.major = element_line(color = "grey30", size=0.1),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 11), legend.title = element_text(size = 14),
        legend.key = element_rect(fill = NA)) +
  guides(color = guide_legend(override.aes = list(size = 4.5)))

ggsave("PCA_bdisp_habitat.png", width = 12, height = 7)

# 3-2. By sampling year
RPY.corr <- betadisper(RP.Bray, RockPools$Year, type = "median", bias.adjust=TRUE)

# 3-2-1. Coordinate mapping*
RPY.corr.crs <- data.frame(RPY.corr$vectors)

RPY.corr.crs$Year <- RockPools$Year  ## Sampling year column "fill-in"

# 3-2-2. PCA with Sampling year coloured
ggplot(RPY.corr.crs, aes(x=PCoA1, y=PCoA2, color=Year)) + geom_point() +
  theme(legend.key = element_rect(fill = NA),
        legend.text = element_text(size = 11),legend.title = element_text(size = 14),
        axis.title = element_blank(),
        panel.background = element_rect(fill = "grey97", colour = "black", size = 1),
        panel.grid.major = element_line(color = "grey86", size=0.1),
        panel.grid.minor = element_blank()) +
  guides(color = guide_legend(override.aes = list(size = 3)))

ggsave("PCA_bdisp_sampling_year.png", width = 10, height = 7)
```

Notice that in fact we don't need to map the coordinates twice. Show that

    RPH.corr <- betadisper(RP.Bray, RockPools$Habitat, type = "median", bias.adjust=TRUE)
    RPH.corr.crs <- data.frame(RPH.corr$vectors)
    RPH.corr.crs$Habitat <- RockPools$Habitat
    RPH.corr.crs$Year <- RockPools$Year

...then change the colour by group in the following ggplot2 commands would produce at least very similar plots.


```{r}
# 4. Centroid vs median
# 4-1. By habitat
# 4-1-1. Plotting the dispersion (From 2-1-5) with bias correction this time
plot(RPH.corr, main=element_blank()) ## Default = Hull

# 4-1-2. Dispersion plot with ellipses, defined as group centroid (Fig. 4a)
plot(RPH.corr, hull = FALSE, ellipse = TRUE, main=element_blank())

# 4-1-3. Similarly, revised 2-1-6 with bias correction (Fig. 4c)
boxplot(RPH.corr, xlab=element_blank())

# 4-2. By sampling year
# 4-2-1. Plotting the dispersion with bias correction (Fig. 4b)
plot(RPY.corr, main=element_blank())

# 4-2-2. Dispersion plot with ellipses, defined as group centroid
plot(RPY.corr, hull = FALSE, ellipse = TRUE, main=element_blank())

# 4-2-3. Similarly, revised 2-2-6 with bias correction (Fig. 4d)
boxplot(RPY.corr, xlab=element_blank())
```
```{r}
# 5. Convex hull plots in ggplot2
# 5-1. By habitat
# 5-1-1. Define convex hull of each habitat group and store into an object

hull.RPH <- RPH.corr.crs %>%
    group_by(Habitat) %>%
    slice(chull(PCoA1, PCoA2))

# 5-1-2. Create a basis: PCA(1,2) of abundance data grouped by habitats

Basis.H <- ggplot(RPH.corr.crs, aes(x=PCoA1, y=PCoA2, color=Habitat)) +
  geom_point(size=0.15) +
  theme(panel.background = element_rect(fill = "grey20", colour = "black", size = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11),
        legend.key = element_rect(fill = NA))

# 5-1-3. Layer the convex hull defined in 5-1
Basis.H.Hull <- Basis.H + aes(fill = Habitat) + geom_polygon(data = hull.RPH, alpha=0.1, size=0.8)   

Basis.H.Hull ## Basis+hull
```

```{r}
# 5-2. By sampling year
# 5-2-1. Define convex hull of each sampling year and store into an object

hull.RPY <- RPY.corr.crs %>%
    group_by(Year) %>%
    slice(chull(PCoA1, PCoA2))

# 5-2-2. Create a basis: PCA(1,2) of abundance data grouped by sampling year

RPY.corr.crs$Year <- as.factor(RPY.corr.crs$Year)

Basis.Y <- ggplot(RPY.corr.crs, aes(x=PCoA1, y=PCoA2, color=Year)) +
  geom_point(size=0.7) +
  theme(panel.background = element_rect(fill = "grey20", colour = "black", size = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11),
        legend.key = element_rect(fill = NA)) +
  labs(fill="Year")

# 5-2-3. Layer the convex hull defined in 5-1
Basis.Y.Hull <- Basis.Y + aes(colour = Year) + geom_polygon(data = hull.RPY, alpha=0.1, size=0.8)

Basis.Y.Hull ## Basis+hull
```

```{r}
# 6. Confidence ellipses plot
# 6-1. By habitat

Basis.H + aes(fill = Habitat) + stat_ellipse(level=0.95, size=1.2) +
  theme(panel.background = element_rect(fill = "grey16", colour = "black", size = 1))

ggsave("Ellipse_habitat.png", width = 10, height = 7)

# 6-2. By sampling year

Basis.Y + aes(colour = Year) + stat_ellipse(level=0.95, size=1.2) +
  theme(panel.background = element_rect(fill = "grey16", colour = "black", size = 1))

ggsave("Ellipse_sampling_year.png", width = 10, height = 7)
```

